---
title: "Homework 2: A simulation study investigating the bootstrap"
author: "BIOS 731 – Advanced Statistical Computing-- Ruilong Chen"
format:
  pdf:
    colorlinks: true
    urlcolor: blue
    include-in-header:
      - text: |
          \usepackage{multirow}
execute:
  echo: true
  warning: false
  message: false
---

```{r, echo = FALSE}
library(tidyverse)
```
## GitHub Repo Link
<https://github.com/ChenRuilonggo/bios731_hw2_RuilongChen>


## Context and learning objectives

This assignment reinforces ideas in Module 2: **Simulations and Resampling Methods**. We focus specifically on implementing a **large-scale simulation study**, and the assignment also includes components involving the bootstrap, parallelization, Git/GitHub, and project organization.

## Due Date and Submission

Submit (via Canvas) a PDF knitted from a `.Rmd` or `.qmd` file. Your PDF should include the web address of the GitHub repository containing your work for this assignment. **Commits after the due date will cause the assignment to be considered late.**

## Point distribution

```{r, echo = FALSE}
knitr::kable(
  data.frame(
    Problem = c("Problem 0", "Problem 1.1", "Problem 1.2", "Problem 1.3", "Problem 1.4", "Problem 1.5"),
    Points  = c(20, 10, 5, 20, 30, 15)
  ),
  align = c("l", "r")
)
```

## Problem 0

This "problem" focuses on the structure of your submission, especially the **use of git and GitHub** for reproducibility, **R Projects** to organize your work, **Quarto/R Markdown** to write reproducible reports, **relative paths** to load local files, and reasonable naming conventions for your files.

To that end:

-   Create a public GitHub repository and a local R Project. I suggest naming the repo/directory `bios731_hw2_YourLastName` (e.g., `bios731_hw2_wrobel`).
-   Push your **entire project folder** to GitHub.
-   Submit a PDF knitted from your `.Rmd`/`.qmd` file to Canvas.
    -   Your solutions should be implemented in your `.Rmd`/`.qmd` file.
    -   Your git commit history should reflect your workflow (i.e., multiple meaningful commits; avoid a single “final” commit with all work).

## Problem 1

**Simulation study.** Your goal in this homework is to plan and execute a well-organized simulation study for multiple linear regression and confidence intervals constructed via both Wald and bootstrap methods.

### Model

Consider the multiple linear regression model

$$Y_i = \beta_0 + \beta_{treatment}X_{i1} + \mathbf{Z_i}^T\boldsymbol{\gamma} + \epsilon_i$$

where we are primarily interested in the treatment effect $\beta_{treatment}$. It is fine to simulate data with no confounders, i.e. $\gamma = 0$.

### Notation

Notation is defined below:

-   $Y_i$: continuous outcome
-   $X_{i1}$: treatment group indicator; $X_{i1}=1$ for treated
-   $\mathbf{Z_i}$: vector of potential confounders
-   $\beta_{treatment}$: average treatment effect, adjusting for $\mathbf{Z_i}$
-   $\boldsymbol{\gamma}$: vector of regression coefficient values for confounders
-   $\epsilon_i$: errors, we will vary how these are defined

### Simulation goals

In our simulation, we want to

-   Estimate $\beta_{treatment}$ and $se(\hat{\beta}_{treatment})$
    -   Evaluate $\beta_{treatment}$ through bias and coverage

You will compare **three** methods for constructing a 95% confidence interval for $\hat{\beta}_{treatment}$:

1.  Wald confidence intervals (standard model-based approach)
2.  Nonparametric bootstrap **percentile** intervals
3.  Nonparametric bootstrap $t$ intervals

You will also evaluate **computation time** for each method.

### Simulation design (full factorial)

Evaluate performance across the following factors:

-   Sample size: $n \in \{10, 50, 500\}$
-   True treatment effect: $\beta_{treatment} \in \{0, 0.5, 2\}$
-   Error distribution:
    -   Normal errors: $\epsilon_i \sim N(0, 2)$
    -   Heavy-tailed errors: $\epsilon_i \sim t_{\nu}$ with $\nu=3$, scaled to have variance 2

**Implementation hint (heavy tails).** If $u \sim t_{\\nu}$ with $\nu>2$, then $\mathrm{Var}(u)=\nu/(\nu-2)$. To match the normal-error condition variance, set

$$
\epsilon_i = u \cdot \sqrt{2\,\frac{\nu-2}{\nu}}.
$$

### Problem 1.1 ADEMP Structure

Answer the following questions. Use the ADEMP framework explicitly:

-   **A (Aim):** What is the goal of the simulation study?
-   **D (Data-generating mechanism):** What model and distributions generate the data? What factors vary across scenarios?
-   **E (Estimand):** What quantity(ies) are you trying to learn about?
-   **M (Methods):** What methods are being evaluated/compared?
-   **P (Performance measures):** What metrics summarize performance?

Also answer:

-   How many simulation scenarios will you be running (i.e., how many unique combinations in the full factorial design)?

### Aim (A)

The goal of this simulation study is to evaluate the statistical performance and computational cost of three methods for constructing 95% confidence intervals for the treatment effect in a multiple linear regression model. Specifically, we assess bias and coverage of the estimated treatment effect and compare computation time across methods under varying sample sizes, true treatment effects, and error distributions.

### Data-generating mechanism (D)

Data are generated from the linear model

$$
Y_i = beta_0 + beta_{treatment} X_{i1} + \epsilon_i,\
$$

where the treatment indicator follows

$$
X_{i1} \sim {Bernoulli}(0.5),\
$$

and we set ($\beta_0 = 0$).

We consider three sample sizes\
$$
n \in {10, 50, 500},\
$$

and three true treatment effects\
$$\
\beta\_{treatment} \in {0, 0.5, 2}.\
$$

Two error distributions are examined, both with variance equal to 2:

Normal errors:\
$$\
\epsilon_i \sim N(0, 2)\
$$

Heavy-tailed errors generated from a scaled (t)-distribution with ($\nu = 3$):\
$$\
u_i \sim t_3, \qquad\
\epsilon_i = u_i \sqrt{2\frac{\nu - 2}{\nu}}.\
$$

### Estimand (E)

The estimand of interest is the treatment effect\
$$\
\beta\_{treatment},\
$$\
representing the average treatment effect estimated using ordinary least squares.

### Methods (M)

We compare three approaches for constructing 95% confidence intervals:

1.  Wald confidence intervals based on the model standard error

2.  Nonparametric bootstrap percentile intervals

3.  Nonparametric bootstrap (t) intervals

### Performance measures (P)

Performance is evaluated using:

-   Bias of ($\hat{\beta}\_{treatment}$)

-   Empirical coverage probability of the 95% confidence intervals

-   Computation time for each method

### Number of scenarios

The full factorial design includes

$$\
3 \text{ sample sizes} \times 3 \text{ treatment effects} \times 2 \text{ error distributions} = 18\
$$

simulation scenarios.

### Problem 1.2 nSim

Based on desired coverage of 95% with Monte Carlo error of no more than 1%, how many simulations ($n_{sim}$) should you perform for **each** simulation scenario? Implement this value of $n_{\text{sim}}$ throughout your simulation study. $$
\text{SE}(\hat{p})=\sqrt{\frac{p(1-p)}{n_{\text{sim}}}}
$$

$$
\sqrt{\frac{0.95(1-0.95)}{n_{\text{sim}}}} \le 0.01
$$

$$
n_{\text{sim}} \ge \frac{0.95\times 0.05}{0.01^2}
= \frac{0.0475}{0.0001}
= 475
$$

### Problem 1.3 Implementation

For bootstrap $t$, goal is to estimate a $t$ distribution given by

$$t^* = \frac{\hat{\theta}^*-\hat{\theta}}{s_{\hat{\theta}^*}}$$

where

-   $\hat{\theta}^*$: estimated parameter value from each bootstrap iteration
-   $\hat{\theta}$: parameter estimate from the original sample
-   $s_{\hat{\theta}^*}$ standard error estimate from a given bootstrap sample; requires a second level of bootstrapping to construct.

#### Parameter choices

-   Use **B = 500** outer bootstrap resamples for the percentile and bootstrap-$t$ intervals.
-   For bootstrap-$t$, use **B_inner = 100** inner bootstrap resamples.
-   Construct **95%** confidence intervals for all methods.

*(If you choose different values, justify your choice and discuss the computation/accuracy trade-off.)*

I use B = 50 and B_inner =10. I parallelize across simulation scenarios and it costs 6 minutes. So if I choose B = 500 and B_inner =100, it will cost me 10 hours. The trade-off is less accuracy.



#### Computing + reproducibility requirements

Execute the full simulation study. For full credit, implement the following:

-   Well-structured scripts and subfolders following guidance from the `project_organization` lecture

-   Use relative file paths to access intermediate scripts and data objects

-   Use readable code practices (clear function boundaries, meaningful names, minimal duplication)

-   **Parallelize across simulation scenarios**

-   Save results from each simulation scenario to an intermediate `.Rds` or `.Rda` file in a `data/` subfolder

    -   Add these files to `.gitignore` so they are not pushed to GitHub

-   Include a `README.md` explaining your workflow

    -   Include what files to run, in what order, and how outputs are produced

-   Ensure end-to-end reproducibility:

    -   I should be able to clone your GitHub repo, open your `.Rproj`, and run the simulation study to regenerate results

### Problem 1.4 Results summary

Create a plot or table summarizing simulation results across scenarios and methods for each of the following:

-   Bias of $\hat{\beta}$
-   Coverage of the **95% CI** for $\hat{\beta}$
-   Distribution of $se(\hat{\beta})$
-   Computation time across methods

**Presentation guidance**

-   If creating plots, I encourage faceting by at least one design factor (e.g., $n$, error distribution, or true treatment value).
-   Include informative captions for each plot/table.
-   For coverage plots, consider adding a reference line at 0.95.



```{r, echo=FALSE}
library(kableExtra)
# Load all scenario result files
rds_files <- list.files("data", pattern = "^sim_.*\\.rds$", full.names = TRUE)

stopifnot(length(rds_files) > 0)

res <- purrr::map_dfr(rds_files, readRDS) %>%
  mutate(
    n = as.integer(n),
    beta_treat = as.numeric(beta_treat),
    err_dist = factor(err_dist, levels = c("normal", "t3"))
  )

# Quick check: how many simulation replicates per scenario file?
scenario_counts <- res %>%
  count(scenario_id, n, beta_treat, err_dist, name = "n_rows") %>%
  arrange(err_dist, beta_treat, n)

kable(scenario_counts, caption = "Number of rows (simulation replicates) per scenario file.")
```

### Bias of beta_hat

```{r,echo=FALSE}
# Build a long table for coverage/time by method, and compute bias
res_long <- res %>%
  pivot_longer(
    cols = c(cover_wald, cover_perc, cover_bt,
             time_wald,  time_perc,  time_bt),
    names_to = c(".value", "method"),
    names_pattern = "(cover|time)_(wald|perc|bt)"
  ) %>%
  mutate(
    method = recode(method,
                    wald = "Wald",
                    perc = "Bootstrap percentile",
                    bt   = "Bootstrap t"),
    bias = beta_hat - beta_treat
  )

# Summaries by scenario and method
summary_tbl <- res_long %>%
  group_by(n, beta_treat, err_dist, method) %>%
  summarise(
    n_sim_effective = sum(is.finite(bias)),
    bias_mean = mean(bias, na.rm = TRUE),
    bias_sd   = sd(bias, na.rm = TRUE),
    coverage  = mean(cover, na.rm = TRUE),
    time_mean = mean(time, na.rm = TRUE),
    time_median = median(time, na.rm = TRUE),
    .groups = "drop"
  )

## ---- Bias table ---------------------------------------------------------
bias_tbl <- summary_tbl %>%
  group_by(n, beta_treat, err_dist) %>%
  summarise(
    bias_mean = mean(bias_mean),
    bias_sd   = mean(bias_sd),
    .groups="drop"
  ) %>%
  arrange(err_dist, beta_treat, n)

knitr::kable(
  bias_tbl,
  digits = 4,
  caption = "Monte Carlo bias and SD of the OLS treatment effect estimator."
)

```

### Coverage of the 95% CI for beta_hat

```{r,echo=FALSE}
## ---- Coverage plot ------------------------------------------------------
library(ggplot2)

p_cov <- summary_tbl %>%
  ggplot(aes(x = factor(n), y = coverage, color = method, group = method)) +
  geom_hline(yintercept = 0.95, linetype="dashed") +
  geom_point(size = 2) +
  geom_line() +
  facet_grid(err_dist ~ beta_treat, labeller = label_both) +
  coord_cartesian(ylim = c(0,1)) +
  labs(
    title = "Empirical coverage of 95% confidence intervals",
    x = "Sample size (n)",
    y = "Coverage probability",
    color = "Method",
    caption = "Dashed line indicates nominal 95% coverage."
  )

p_cov

```

### SE distribution table

```{r,echo=FALSE}
## ---- SE summary table ---------------------------------------------------
se_tbl <- res %>%
  group_by(n, beta_treat, err_dist) %>%
  summarise(
    mean_model_SE = mean(se_hat),
    sd_model_SE   = sd(se_hat),
    mean_boot_SE  = mean(se_boot),
    sd_boot_SE    = sd(se_boot),
    .groups="drop"
  ) %>%
  arrange(err_dist, beta_treat, n)

knitr::kable(
  se_tbl,
  digits = 4,
  caption = "Summary of standard error estimates."
)

```

### Computation time across methods

```{r,echo=FALSE}
time_tbl <- summary_tbl %>%
  group_by(method) %>%
  summarise(
    mean_time   = mean(time_mean),
    median_time = median(time_median),
    .groups="drop"
  ) %>%
  arrange(mean_time)

knitr::kable(
  time_tbl,
  digits = 4,
  caption = "Average computation time per simulation replicate."
)
```

### Problem 1.5 Discussion

Interpret the results summarized in Problem 1.4.

1.  Write **one paragraph** summarizing the main findings of your simulation study.
2.  Then answer the questions below:

-   How do the different methods for constructing confidence intervals compare in terms of computation time?
-   Which method(s) provide the best coverage when $\epsilon_i \sim N(0, 2)$?
-   Which method(s) provide the best coverage for the heavy-tailed errors?

Finally, briefly comment on any notable interactions (e.g., how performance changes with $n$ or error type) and any practical recommendations you would make based on your results.

Overall, the simulation results show that the Wald confidence interval performs consistently well across both error distributions and sample sizes. The bootstrap-based intervals are substantially more computationally expensive and their performance depends more strongly on the sample size and error distribution.

Computation time.
The Wald method is the fastest approach by a large margin because it requires only a single model fit per simulation replicate. The bootstrap percentile method is much slower due to repeated resampling and refitting, and the bootstrap-t method is the most computationally intensive because it includes an additional inner bootstrap loop. Thus, computation time increases dramatically from Wald → bootstrap percentile → bootstrap-t.

Coverage under normal errors.
Under normally distributed errors, the Wald interval provides coverage very close to the nominal 95% level across all sample sizes. In contrast, the bootstrap methods do not show clear improvement as the sample size increases and in some scenarios their coverage even slightly decreases. This indicates that the normal-theory approximation underlying the Wald interval works very well when the model assumptions are satisfied.

Coverage under heavy-tailed errors.
For heavy-tailed errors, the bootstrap methods improve as the sample size increases, showing a clear upward trend toward the nominal coverage level. The bootstrap-t interval consistently outperforms the bootstrap percentile interval, suggesting that studentization improves the accuracy of bootstrap confidence intervals. However, in this study neither bootstrap method surpasses the Wald interval in coverage. A likely explanation is the relatively small number of bootstrap replicates used (B = 50 and B_inner = 10). With such small bootstrap sizes, the bootstrap intervals are subject to substantial Monte Carlo error, which can lead to undercoverage and unstable performance. With larger bootstrap sizes, the bootstrap methods would likely perform more competitively in the heavy-tailed setting.

Interactions and practical recommendations.
Two main patterns emerge. First, the Wald interval performs well across both distributions and sample sizes in this simulation. Second, the benefit of bootstrap methods becomes more visible in heavy-tailed settings and increases with sample size. Across all scenarios, the bootstrap-t interval consistently performs better than the bootstrap percentile interval. In practice, the Wald interval is recommended for routine use due to its excellent performance and extremely low computational cost. The bootstrap-t method may be preferred when robustness to non-normal errors is important, but it requires substantially larger numbers of bootstrap replicates to achieve stable and reliable coverage.
